The Dynamics of Anthropomorphic Effects in Human-Robot Interaction

### 2014-03-25 paper.tex
new bib item and anthropomorphic design section almost done

The Uncanny Valley hypothesis \citep{mori_uncanny_1970} proposes that human-like
designs of objects and non-human agents evoke a positive feeling of affinity in
the human observer / user, which in turn reflects the experience of emotional
engagement with and feelings of empathy for the non-human agent
\citep{cheetham_human_2011}\fxfatal{uncanny valley is not about positive
feeling!}. The positive effect of so-called \textbf{anthropomorphic form} is
exploited in the design of artificial agents
(\eg digital characters, robots) for several reasons. Amongst others,
anthropomorphic design should serve to evoke a positive valence as the Uncanny
Valley hypothesis suggests, which in turn should foster people's acceptance of
the artificial agent, and enable effective (human-like and especially
human-social) interaction.  However, at a certain point as the non-human agent
approaches but fails to closely resemble a human, the evoked valence in the
human user suddenly turns into a negative uncanny feeling of what
\cite{mori_uncanny_1970} describes as disgust and revulsion. This, in turn,
might lead to rejection of the non-human but human-like agent.  The Uncanny
Valley hypothesis predicts that with further increasing the human-likeness of
the non-human agent such that it appears so realistic that it is not
distinguishable from a human but perceived to be human, the effect is again a
positive one, reflecting feelings and engagement that are similar as those
evoked by a human-human interaction.  Of course, roboticists and developers try
to avoid ``falling'' into this Uncanny Valley with their human-like designs of
artificial agents. Consequently, there is a considerable research effort to
better understand the boundaries of anthropomorphic design and their effects
on the interaction and experience. To date, findings have been however
inconsistent \citep{cheetham_human_2011}.

@@

An artifact-centered perspective explains anthropomorphism from the ascertainable 
parts / the design of the non-human agent.
The direction of causality in this view is that the non-human artifact (or a certain 
feature of it) encourages the human to anthropomorphize the artifact / agent.
This perspective assumes that humans directly respond to life-like and social cues 
that the non-human agent or system emits, without thoughtful mental processing, 
by simply applying stereotypes and heuristics to it. Humans naturally react to specific 
cues. This is natural because, for instance, from early childhood 
on, humans are inherently well-trained to perceive life \citep{epley_seeing_2007}, 
or to respond to social cues, such as facial expressions, social gestures or turn-
taking. 
These cues can be of various senses and modalities. One of the most important aspects 
is the outer appearance of the unknown agent or artifact. \cite{schmitz_concepts_2011} 
argues that an agent's outer appearance is likely to have an important impact on how 
the object is overall perceived (how it appears to the human). Thus, if a robot looks much like a human, it appears to be like a human, and is perceived accordingly.

This fact is exploited in the design of robots and other things. Again, the basic 
assumption here is that if an artifact 
somehow resembles a human, it is likely to be treated similar to a human. If this 
explanation of anthropomorphism is correct, people may respond automatically also to 
social cues emitted by an artificial embodied agent such as a robot, and apply 
human-human social schemas and norms to the occurring interactions. \hl{(However, at 
the same time, the human-like design of things also raises expectations from the 
human user side, to which we will come back later in more detail.)}
The human-like design of robots implies various consequences, of which we will present some in the following.

\subsubsection{Anthropomorphic design of robots and the Uncanny Valley hypothesis\\}

%People's tendency to anthropomorphize technologies has been described in HCI 
%\citep{reeves_media_1996} and related to digital agents; however the phenomenon 
%seems to be more coined in HRI. There are several reasons for this, one of it stems 
%from the unique characteristics of robots (their typical design).

What do we mean by \textit{human-like / anthropomorphic form}?
Using the term \textit{``form''} to describe the design of an artifact, we refer 
not only to the shape of an artifact (how the artifact looks like) but to the total 
expression of the product \citep{bartneck_shaping_2004}. \cite{disalvo_hug:_2003} 
argues that \textit{form} includes a product's physical shape, materials, and 
behavioral qualities. Latter aspect is of special importance for interactive 
systems that show some degree of autonomy (independence).
Several categories of anthropomorphic form can be classified, but there is no 
common agreement on the borders and transitions between these categories: 
anthropomorphic, zoomorphic, caricatured, functional \citep{fong_survey_2003}.

The use of anthropomorphic (human-like) form in the design of non-human 
things is as old as people's tendency to anthropomorphize. Anthropomorphic (and 
also animal-shaped) forms were used already in prehistoric artworks and 
illustrations, and they have a long-standing tradition in the design of daily life 
artifacts and in general \citep{guthrie_bottles_2000}. The principle of human-like 
form is probably most prominent in the design of robots, due to several reasons. 
However, it has been controversially discussed since the early 1970ies, related to 
\cite{mori_uncanny_1970} theory of the \texit{Uncanny Valley}, see Figure \ref{fig:uncanny_valley}.

The underlying crucial point about the design are the implications that it holds, and the 
suggestions for interactions that it makes to the user. In other words, the appearance 
and function of a product, whether it is an appliance or a humanoid robot, affects the 
way that people perceive it, interact with it, and build long-term relationships with 
it \cite{bartneck_shaping_2004}. The relation between the design of a robot and the 
user's perception of it, is also formulated in the \textit{Uncanny Valley} hypothesis \citep{mori_uncanny_1970}.

@@

Most work that studies how far human users treat robots as social agents, builds on the 
famous work of \cite{reeves_media_1996} in HCI. In the so-called \textit{Media 
Equation}, the authors describe that people mindlessly react in a social way to their computer 
and that they tend to even attribute human (social) characteristics to it -- despite 
the fact that a computer does not have human-like characteristics. Apart from 
computers, anthropomorphism has also been observed with digital / virtual characters 
\citep{reeves_media_1996,nass_anthropocentrism_1995}, and artificial embodied agents, 
such as robots. In short, one can say that with robots, and human-like robots in 
particular, anthropomorphism is a commonly observed phenomenon. However, the 
anthropomorphic design approach in robotics is controversially discussed.

But what are the motivations which underlie the efforts of enhancing anthropomorphism through specific anthropomorphic design in robots? Why are anthropomorphic forms so distinct and at the same time controversial in robotics? There is no one single answer to this question.

\hl{human-likeness has a positive effect\\}

On one hand, developers and designers have discovered that human-like robots may better match the affordances that our to humans adapted world brings along. Also. anthropomorphic form and people's resulting consequence to anthropomorphize, appears to be generally beneficial for human-robot relations, especially with socially interactive robots \cite{fong_survey_2003}. Consequently, the domain of (social) robotics tries to exploit this fact by enhancing anthropomorphism through anthropomorphic robots: \textit{"Applying anthropomorphism or zoomorphism changes the way how users try to understand and make sense of a [system] by projecting everyday \hl{expectations} of human and animalistic life onto it."} \cite{schmitz_concepts_2011}.

@@

\hl{human-likeness has a negative effect\\}

However, the anthropomorphic form also 
increases people's expectations of the agent, and consequently, a very human-like 
appearing robot is expected to act like a human. But today's artificial systems are 
not able (yet) to mimic human movements, or human social cues in an acceptable way. 
The case is that the artificial agent pretends to be human but fails, and this is 
more than just disappointing for the 
human -- it appears uncanny and evokes reluctance, the so-called 
\textit{"uncanny valley"}. The idea of naming this shift and drop from a positive to a negative reaction,
follows Freud's description of the so-called \textit{uncanny} (a translation from the German word 
\textit{``unheimlich''}). Controversially, something uncanny \textit{"derives its terror not from 
something externally alien or unknown but -- on the contrary -- from something 
strangely familiar which defeats our efforts to separate ourselves from it"} (cited 
after \cite{hegel_understanding_2008}).
In other words, humans are very sensitive to the limits of what is familiar to them 
in a sense of being human. At the same time, humans are naturally very skilled when 
it comes to identifying what is acceptable as a human and what is an imitation 
of it. Even subtle differences can be crucial -- humans are inherently well trained 
in identifying if there is \hl{``something strange''} in the familiar.
Summing it up, a not yet very well defined thin line separates acceptable human-like forms in 
robots from creepy artificial \textit{almost} human-like appearing robots.

@@

\hl{(ethical issues, maybe put this rather in the end)}
Another critical point with the use of human-like form in artificial agents are 
ethical aspects. For instance, we need to consider how far we actually want an 
\textit{artificial agent} to resemble a human -- or ourself. Even though we do not 
discuss these aspects in this article, we think the HRI / robotics community needs 
to consider questions of how far an artificial agent should act ``as if it would be 
social'' toward humans. What implications does an anthropomorphic design of robots 
have? What happens as soon as the artificial agent is considered a \textit{moral 
agent} \cite{sullins_when_2006}? Using \textit{socially interactive robots} in 
contexts involving elderly care or rehabilitation for disabled / handicapped 
persons, autistic children is a controversially discussed topic 
\cite{robins_robots_2005}.

@@

%For instance, recent studies found that people tend to ascribe personality and 
%intentions to robots \hl{(REFS)} and that they show empathetic behavior toward robots 
%\citep{rosenthal-vonderputten_experimental_2013}. Summing it up, there is a huge and 
%constantly growing body of literature that describes anthropomorphism in HCI and HRI. 
%Despite the phenomenon is well studied, we still know rather little about it, so that 
%at conferences and symposiums, anthropomorphism is still a ``hot topic''; however, at 
%the same time theories related to anthropomorphism, such as the \textit{``Uncanny 
%Valley''} are controversially discussed.

@@

Anthropomorphism represents just one of many examples of induction whereby "people reason about an unknown stimulus based on a better-known representation of a related stimulus" \cite{epley_when_2008}, in this case reasoning about a non-human agent based on representation of the self or other humans. This supports the usual role we assign to anthropomorphism in robotics: the anthropomorphic design of a robot encourages people to reason about this unfamiliar system \emph{as if} it was a human. To increase this tendency, robots are designed with anthropomorphic forms (human-like or lifelike forms, in the broader sense), or are able to display and use human-social cues when interacting with humans.\\

@@

To illustrate better how anthropomorphism differs from attributions of animacy or life-likeness, it is worth looking at paraphrases and examples of it. That is to say, anthropomorphism can alternatively be understood as \textbf{humanization} or \textbf{personification} of anything other than a human being. Personification has ancient roots, \eg in mythology but also in storytelling. Anthropomorphism was also used to describe and be able to predict natural forces or anticipate great spirits, for instance. An illustration of an anthropomorphism related to nature is shown in Figure \ref{fig:north-wind-sun}. In this Milo Winter (1919) drawing \textit{"The North Wind and the Sun"}, both north wind and sun are given a human face.

@@

Anthropomorphism is used in different senses throughout different disciplines \citep{duffy_anthropomorphism_2003} which makes it difficult to synthesize findings. Depending on which understanding of anthropomorphism is applied, not everything colloquially labeled as \emph{anthropomorphism} really is anthropomorphism.\footnote{For instance, it is not always clear whether a person (may it be an adult or a child) who gives and anthropomorphic response, believes that the other agent really thinks or acts like a human or whether she is using anthropomorphism in a metaphoric sense.\citep{leeds_childrens_1992}} We need to be clear about what we mean by anthropomorphism because at the moment, there are different, even partially conflicting understandings of anthropomorphism, not only across but also within disciplines \cite{duffy_anthropomorphism_2002}. In particular, in the multi-disciplinary field of HRI, anthropomorphism is understood and interpreted in such a variety of ways that it is almost impossible to bring together and compare the research work done. In turn, it seems that the understanding of anthropomorphism in HRI has not improved much over the last decade but has only become more diverse (which is ok, too, but certainly a deeper understanding and common grounds would be convenient).

@@

\paragraph{Attractiveness Delta}

Besides, we propose to call \emph{attractiveness delta}, noted $\Delta_{a}$, the
difference between the initial capital of anthropomorphism \ICA for a given tuple (user,
robot, context) and the anthropomorphism peak \AntMax (Figure~\ref{fig:dynamics}).
This can be interpreted as the initial \emph{excitement} over the robot, or how
much a user tend to increase his/her anthropomorphic engagement when he/she
starts to actually interact with the robot.

@@

@@

For instance, we propose that the human \emph{adapts} iteratively to the robot
by restating its cognitive model of the robot. Could our two models be
conversely relied on to have the robot itself iteratively adapting its
behaviour? In particular, we hypothesize that the so-called novelty effect
represents a peak in anthropomorphic manifestations, followed by mostly
deception-driven refinements of the cognitive behavioural model of the robot.
Could we imagine that the robot pro-actively \emph{verbalises} its own limits,
in a timely manner (likely before the end of the novely effect). And how would
this affect the cognitive models built by the human?

@@

@@

>>>>

The Uncanny Valley hypothesis \citep{mori_uncanny_1970} proposes that human-like
designs of objects and non-human agents can evoke both positive and negative feelings 
of affinity in the human observer / user. Positive feelings reflect the experience of 
emotional engagement with and feelings of empathy for the non-human agent
\citep{cheetham_human_2011}. The positive effect of so-called
\textbf{anthropomorphic form} is exploited in the design of artificial agents
(\eg digital characters, robots) for several reasons. Amongst others,
anthropomorphic design should serve to evoke a positive valence as the Uncanny
Valley hypothesis suggests, which in turn should foster people's acceptance of
the artificial agent, and enable effective (human-like and especially
human-social) interaction.  However, at a certain point as the non-human agent
approaches but fails to closely resemble a human, the evoked valence in the
human user suddenly turns into a negative uncanny feeling of what
\cite{mori_uncanny_1970} describes as disgust and revulsion. This, in turn,
might lead to rejection of the non-human but human-like agent.  The Uncanny
Valley hypothesis predicts that with further increasing the human-likeness of
the non-human agent such that it appears so realistic that it is not
distinguishable from a human but perceived to be human, the effect is again a
positive one, reflecting feelings and engagement that are similar as those
evoked by a human-human interaction.  Of course, roboticists and developers try
to avoid ``falling'' into this Uncanny Valley with their human-like designs of
artificial agents. Consequently, there is a considerable research effort to
better understand the boundaries of anthropomorphic design and their effects
on the interaction and experience. To date, findings have been however
inconsistent \citep{cheetham_human_2011}.

@@

An artifact-centered perspective explains anthropomorphism from the ascertainable 
parts / the form of the non-human agent, namely the anthropomorphic design of it.
Using the term \textit{``form''} to describe the design of an artifact, we refer 
not only to the shape of an artifact (how the artifact looks like) but to the total 
expression of the product \citep{bartneck_shaping_2004}. \cite{disalvo_hug:_2003} 
argues that \textit{form} includes a product's physical shape, materials, and 
behavioral qualities. Latter aspect is of special importance for interactive 
systems that show some degree of autonomy (independence).
Several categories of anthropomorphic form can be classified, but there is no 
common agreement on the borders and transitions between these categories: 
anthropomorphic, zoomorphic, caricatured, functional \citep{fong_survey_2003}.
The use of nature-inspired and in particular anthropomorphic (human-like) form in the 
design of non-human things is as old and manifold as people's tendency to anthropomorphize. 
Anthropomorphic (and also animal-shaped) forms were used already in prehistoric 
artworks and illustrations, and they have a long-standing tradition in design in 
general, and with respect of the design of daily life things, in particular 
\citep{norman_design_1988,guthrie_bottles_2000}.

The direction of causality in the artifact-centered view is that the non-human artifact 
(or a certain feature of it) encourages the human to anthropomorphize the artifact / 
agent. This perspective assumes that humans directly respond to life-like and social 
cues that the non-human agent or system emits, without thoughtful mental processing, 
by simply applying stereotypes and heuristics to it, and in turn apply 
human-human social schemas and norms to the occurring interactions.
This view is acceptable since humans naturally react to specific 
cues. From early childhood 
on, humans are inherently well-trained to perceive life \citep{epley_seeing_2007}, 
or to respond to social cues, such as facial expressions, social gestures or turn-taking. 
One has to keep in mind that these cues are of various senses and modalities in human-
human interaction, and consequently human-like design also addresses several senses. 
One of the most outstanding aspects is probably the outer appearance of an agent or 
artifact (in particular when it is unknown or unfamiliar). \cite{schmitz_concepts_2011} 
argues that an agent's outer appearance is likely to have an important impact on how 
the object is overall perceived. Consequently, if a robot's outer form resembles a 
human, the robot is likely to be treated similar to a human \hl{(\eg eye-contact is 
established or the robot is greeted)}.

People's natural reaction to anthropomorphic design is exploited in the design of 
robots (in particular, social or also called \textit{sociable} robots 
\citep{breazeal_sociable_2000}) and other things, however, consequences can be both of 
positive and negative nature, which makes anthropomorphic form an interesting subject 
of study.

Most work that studies how far human users treat robots as social agents, builds on the 
famous work of \cite{reeves_media_1996} in HCI. In the so-called \textit{Media 
Equation}, the authors describe that people mindlessly react in a social way to their computer 
and that they tend to even attribute human (social) characteristics to it -- despite 
the fact that a computer does not have human-like characteristics. Apart from 
computers, anthropomorphism has also been observed with digital / virtual characters 
\citep{reeves_media_1996,nass_anthropocentrism_1995}, and artificial embodied agents, 
such as robots. In short, one can say that with robots, and human-like robots in 
particular, anthropomorphism is a commonly observed phenomenon. 
But what are the motivations which underlie the efforts of enhancing anthropomorphism 
through specific anthropomorphic design in robots? Why are anthropomorphic forms so 
distinct and at the same time controversial in robotics? There is no one single answer 
to these questions.

The underlying central point about the human-like design of robots lies in the 
implications that this design holds, and the suggestions for interactions that
it makes to the human user. 
In other words, the appearance 
and function of a product, whether it is an appliance or a humanoid robot, affects how 
people perceive it, interact with it, and build long-term relationships with 
it \cite{bartneck_shaping_2004}. 
The principle of human-like form is probably most prominent in the design of robots, 
due to several reasons, such as to foster efficient interaction and increase 
acceptance. An anthropomorphic design of a robot encourages people to reason about this 
unfamiliar system \emph{as if} it was a human. To increase this tendency, robots are 
designed with anthropomorphic (or lifelike) forms, or are able to display and use 
human-social cues when interacting with humans.
However, anthropomorphic robot designs have been controversially discussed and received 
remarkable attention since the early 1970ies, related to \cite{mori_uncanny_1970} 
theory of the \textit{Uncanny Valley}, see Figure \ref{fig:uncanny_valley}.

@@

\paragraph*{Anthropomorphic form can be advantageous for HRI\\}

On one hand, developers and designers have discovered that human-like robots may better 
match the affordances that our to humans adapted world brings along. Also, up to a 
certain point, anthropomorphic form as it increases anthropomorphism,
appears to be generally beneficial for human-robot relations, especially with socially 
interactive robots \cite{fong_survey_2003}. Consequently, the domain of (social) 
robotics tries to exploit this fact by enhancing anthropomorphism through 
anthropomorphic robots: Applying anthropomorphism or zoomorphism changes the 
way how users try to understand and make sense of a system by projecting everyday 
\hl{expectations} of human and animalistic life onto it. \cite{schmitz_concepts_2011}.

@@

Anthropomorphism represents just one of many examples of induction whereby "people reason about an unknown stimulus based on a better-known representation of a related stimulus" \cite{epley_when_2008}, in this case reasoning about a non-human agent based on representation of the self or other humans.

\paragraph*{Anthropomorphic form can be disadvantageous for HRI\\}

The anthropomorphic form also 
increases people's expectations of the agent, and consequently, a very human-like 
appearing robot is not only perceived but also expected to act like a human. However, today's artificial systems are 
not able (yet) to well mimic human movements, or human social cues in an acceptable way. 
The case is that the artificial agent pretends to be human but fails, and this is 
more than just disappointing for the 
human -- it appears uncanny and evokes reluctance, the so-called 
\textit{"uncanny valley"}. The idea of naming this shift and drop from a positive to a negative reaction,
follows Freud's description of the so-called \textit{uncanny} (a translation from the German word 
\textit{``unheimlich''}). Controversially, something uncanny \textit{"derives its terror not from 
something externally alien or unknown but -- on the contrary -- from something 
strangely familiar which defeats our efforts to separate ourselves from it"} (cited 
after \cite{hegel_understanding_2008}).
In other words, humans are very sensitive to the limits of what is familiar to them 
in a sense of being human. At the same time, humans are naturally very skilled when 
it comes to identifying what is acceptable as a human and what is an imitation 
of it. Even subtle differences can be crucial -- humans are inherently well trained 
in identifying if there is \hl{``something strange''} in the familiar.
Summing it up, a not yet very well defined thin line separates acceptable human-like forms in 
robots from creepy artificial \textit{almost} human-like appearing robots.

@@

@@

%For instance, recent studies found that people tend to ascribe personality and 
%intentions to robots \hl{(REFS)} and that they show empathetic behavior toward robots 
%\citep{rosenthal-vonderputten_experimental_2013}.

@@

@@

To illustrate better how anthropomorphism differs from attributions of animacy or life-
likeness, it is worth looking at paraphrases and examples of it. That is to say, 
anthropomorphism can alternatively be understood as \textbf{humanization} or 
\textbf{personification} of anything other than a human being. Personification has 
ancient roots, \eg in mythology but also in storytelling. Anthropomorphism was also 
used to describe and be able to predict natural forces or anticipate great spirits, for 
instance. An illustration of an anthropomorphism related to nature is shown in Figure 
\ref{fig:north-wind-sun}. In this Milo Winter (1919) drawing \textit{"The North Wind 
and the Sun"}, both north wind and sun are given a human face.

@@

Anthropomorphism is used in different senses throughout different disciplines 
\citep{duffy_anthropomorphism_2003} which makes it difficult to synthesize findings. 
Depending on which understanding of anthropomorphism is applied, not everything 
colloquially labeled as \emph{anthropomorphism} really is anthropomorphism.
\footnote{For instance, it is not always clear whether a person (may it be an adult or 
a child) who gives and anthropomorphic response, believes that the other agent really 
thinks or acts like a human or whether she is using anthropomorphism in a metaphoric 
sense.\citep{leeds_childrens_1992}} We need to be clear about what we mean by 
anthropomorphism because at the moment, there are different, even partially conflicting 
understandings of anthropomorphism, not only across but also within disciplines 
\cite{duffy_anthropomorphism_2002}. In particular, in the multi-disciplinary field of 
HRI, anthropomorphism is understood and interpreted in such a variety of ways that it 
is almost impossible to bring together and compare the research work done. In turn, it 
seems that the understanding of anthropomorphism in HRI has not improved much over the last decade but has only become more diverse (which is ok, too, but certainly a deeper 
understanding and common grounds would be convenient).

@@

@@

\subsection{Limitations}

@@

For instance, we propose that the human \emph{adapts} iteratively to the robot
by restating its cognitive model of the robot. Could our two models be
conversely relied on to have the robot itself iteratively adapting its
behavior? In particular, we hypothesize that the so-called novelty effect
represents a peak in anthropomorphic manifestations, followed by mostly
deception-driven refinements of the cognitive behavioral model of the robot.
Could we imagine that the robot pro-actively \emph{verbalizes} its own limits,
in a timely manner (likely before the end of the novelty effect). And how would
this affect the cognitive models built by the human?

@@

\citep{ruijten_introducing_2014}

@@

\subsection{Ethical aspects of anthropomorphism and human-like form of robots}

% our message kind of is that it would be favorable to sustain anthropomorphism over 
% extended periods of time ... however, this is ethically questionable, when thinking 
% of emotional bonding with an artificial agent, etc. --> we sould elaborate on this a % bit

Another critical point with the use of human-like form in artificial agents are 
ethical aspects. For instance, we need to consider how far we actually want an 
\textit{artificial agent} to resemble a human -- or ourself. Even though we do not 
discuss these aspects in this article, we think the HRI / robotics community needs 
to consider questions of how far an artificial agent should act ``as if it would be 
social'' toward humans. What implications does an anthropomorphic design of robots 
have? What happens as soon as the artificial agent is considered a \textit{moral 
agent} \cite{sullins_when_2006}? Using \textit{socially interactive robots} in 
contexts involving elderly care or rehabilitation for disabled / handicapped 
persons, autistic children is a controversially discussed topic 
\cite{robins_robots_2005}.
